Finally Tagless, Partially Evaluated

### 2008-11-03 tagless/tagless-final.tex
A few small corrections; new link to the code

We emphasize ``typed'' and ``fold'' in the previous sentence.
We use a typed version of \citets{Mogensen-SelfApplicable}
encoding of the recursive type of terms \citep{bohm-automatic}, which
makes folds over terms much easier to write than functions over
terms that are not primitive recursive (or, compositional).
In contrast, \citearound{'s earlier encoding of the sum type of
terms}\citet{JFP-Mogensen} does not privilege folds.
In exchange, we statically express object types in the metalanguage
and prevent both kinds of run-time errors in
\S\ref{tagproblem}, due to evaluating ill-typed or open terms.
Because the new interpreter
uses no universal type or pattern matching, it never gives a
run-time error, and is in fact total.  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \citep{WalidICFP02}
that arises from error checking at run time.

@@

The evaluator above is wired directly into
functions such as |b|, |lam|, and |app|, whose names appear
free in |testf1| above.
\ifshort We \else In the rest of this paper, we \fi
explain how to abstract over these functions' definitions
and apply different folds to the same object language, so as
to process the same term using many other
interpreters: we can
\begin{itemize}
    \item evaluate the term to a value in the metalanguage;
    \item measure the length of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        even call-by-name (CBN) CPS in a call-by-value (CBV) metalanguage,
        so as to isolate the evaluation
        order of the object language from that of the metalanguage.\ifshort
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}\fi
\end{itemize}
We have programmed all our interpreters and examples in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is
available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
to supplement the paper. 
\ifshort For simplicity, main examples in the paper will be in MetaOCaml;
all examples have also been implemented in Haskell.
\else Except for the basic definitions in \S\ref{encoding},
we show our examples in (Meta)OCaml even though some of our
claims are more obvious in Haskell, for consistency and because
MetaOCaml provides convenient, typed staging facilities.
\fi

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
In contrast to \S\ref{intro}, in the rest of the paper 
we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient, but we have also implemented our
approach using de Bruijn indices
(in \S\ref{S:de-Bruijn} and the accompanying source file
|incope-dB.ml|).

@@

Conversely, the type system of the metalanguage checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, the type checker will complain there that the expected type
|Int| does not match the inferred |Bool|.  Similarly, the object term
$\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Both Haskell's and MetaOCaml's type checker also flags syntactically invalid
object terms, such as if we forget |app| somewhere above.

@@

Besides immediate evaluation, we can compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type~$t$
as values of type |('c, |$t$|) code|, where |'c| is the
environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<|$e$|>.|, which quotes the expression~$e$ for
evaluation at a future stage. The \emph{escape} |.~|$e$ must occur
within a bracket and specifies that the expression~$e$ must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.!|$e$ evaluates
the future-stage code value~$e$ by compiling and linking it at run time.
Bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

@@

We start building the partial evaluator anew and
switch to the data type
\begin{code}
type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr
\end{code}
so that a partially evaluated term always contains a dynamic
component and sometimes contains a static component.
The two
alternative constructors of an |option| value, |Some| and |None|,
tag each partially evaluated term to indicate whether its value is
known statically at the present stage.
This tag is not an object type tag: all pattern matching below
is exhaustive. Now that the future-stage component is always available, we
can define the polymorphic function
\begin{code}
let abstr1 (P1 (_,dyn) : ('c,'dv) repr) : ('c,'dv) C.repr = dyn
\end{code}
to extract it without needing polymorphic lift into~|C|.  We then try
to define the term combinators\nobreak\hspace{0pt}---and get as far as
the first-order constructs of our object language, including |if_|.
\begin{code}
let int (x:int) = P1 (Some (R.int x), C.int x)
let add e1 e2 = match (e1,e2) with
  | (P1 (Some n1, _), P1 (Some n2, _)) -> int (R.add n1 n2)
  | _ -> P1 (None, C.add (abstr1 e1) (abstr1 e2))
let if_ eb et ee = match eb with
  | P1 (Some s, _) -> if s then et () else ee ()
  | _ -> P1 (None, C.if_ (abstr1 eb) (fun () -> abstr1 (et ()))
                                     (fun () -> abstr1 (ee ())))
\end{code}
However, we stumble on functions.  Given how we just
defined~|repr|, a partially evaluated object function, such as the
identity $\fun{x}x$ (of type $\ZZ\to\ZZ$) embedded in OCaml as
|lam (fun x -> x)| (of type |('c,int->int) repr|), consists of
a dynamic part (of type |('c,int->int) C.repr|) and optionally
a static part (of type |('c,int->int) R.repr|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

>>>>

We emphasize ``typed'' and ``fold'' in the previous sentence.
We use a typed version of \citets{Mogensen-SelfApplicable}
encoding of the recursive type of terms \citep{bohm-automatic}, which
makes it much easier to write folds over terms than term functions
that are not primitive recursive (or, compositional).
In contrast, \citearound{'s earlier encoding of the sum type of
terms}\citet{JFP-Mogensen} does not privilege folds.
In exchange, we statically express object types in the metalanguage
and prevent both kinds of run-time errors in
\S\ref{tagproblem}, due to evaluating ill-typed or open terms.
Because the new interpreter
uses no universal type or pattern matching, it never gives a
run-time error, and is in fact total.  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \citep{WalidICFP02}
that arises from error checking at run time.

@@

The evaluator above is wired directly into
functions such as |b|, |lam|, and |app|, whose names appear
free in |testf1| above.
\ifshort We \else In the rest of this paper, we \fi
explain how to abstract over these functions' definitions
and apply different folds to the same object language, so as
to process the same term using many other
interpreters: we can
\begin{itemize}
    \item evaluate the term to a value in the metalanguage;
    \item measure the length of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        even call-by-name (CBN) CPS in a call-by-value (CBV) metalanguage,
        so as to isolate the evaluation
        order of the object language from that of the metalanguage.\ifshort
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}\fi
\end{itemize}
We have programmed all our interpreters and examples in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is
available at \url{http://okmij.org/ftp/tagless-final/}
to supplement the paper. 
\ifshort For simplicity, main examples in the paper will be in MetaOCaml;
all examples have also been implemented in Haskell.
\else Except for the basic definitions in \S\ref{encoding},
we show our examples in (Meta)OCaml even though some of our
claims are more obvious in Haskell, for consistency and because
MetaOCaml provides convenient, typed staging facilities.
\fi

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
In contrast to \S\ref{intro}, in the rest of the paper 
we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensure that our object programs are closed.
We find HOAS to be more convenient, but we have also implemented our
approach using de Bruijn indices
(in \S\ref{S:de-Bruijn} and the accompanying source file
|incope-dB.ml|).

@@

Conversely, the type system of the metalanguage checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, the type checker will complain there that the expected type
|Int| does not match the inferred |Bool|.  Similarly, the object term
$\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Both Haskell's and MetaOCaml's type checkers also flag syntactically invalid
object terms, such as if we forget |app| somewhere above.

@@

Besides immediate evaluation, we can compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type~$t$
as values of type |('c, |$t$|) code|, where |'c| is the
environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<|$e$|>.|, which quotes the expression~$e$ for
evaluation at a future stage. The \emph{escape} |.~|$e$ must occur
within a bracket and specifies that the expression~$e$ must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.!|$e$ evaluates
the future-stage code value~$e$ by compiling and linking it at run time.
Bracket, escape, and run are akin (modulo hygiene) to
quasi-quotation, unquotation, and |eval| of Lisp.

@@

We start building the partial evaluator anew and
switch to the data type
\begin{code}
type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr
\end{code}
so that a partially evaluated term always contains a dynamic
component and sometimes contains a static component.
The two
alternative constructors of an |option| value, |Some| and |None|,
tag each partially evaluated term to indicate whether its value is
known statically at the present stage.
This tag is not an object type tag: all pattern matching below
is exhaustive. Now that the future-stage component is always available, we
can define the polymorphic function
\begin{code}
let abstr1 (P1 (_,dyn) : ('c,'dv) repr) : ('c,'dv) C.repr = dyn
\end{code}
to extract it without needing polymorphic lift into~|C|.  We then try
to define the term combinators\nobreak\hspace{0pt}---and get as far as
the first-order constructs of our object language, including |if_|.
\begin{code}
let int (x:int) = P1 (Some (R.int x), C.int x)
let add e1 e2 = match (e1,e2) with
  | (P1 (Some n1, _), P1 (Some n2, _)) -> int (R.add n1 n2)
  | _ -> P1 (None, C.add (abstr1 e1) (abstr1 e2))
let if_ eb et ee = match eb with
  | P1 (Some s, _) -> if s then et () else ee ()
  | _ -> P1 (None, C.if_ (abstr1 eb) (fun () -> abstr1 (et ()))
                                     (fun () -> abstr1 (ee ())))
\end{code}
However, we stumble on functions.  Given how we just
defined~|repr|, a partially evaluated object function, such as the
identity $\fun{x}x$ (of type $\ZZ\to\ZZ$) embedded in OCaml as
|lam (fun x -> x)| (of type |('c,int->int) repr|), consists of
a dynamic part (of type |('c,int->int) C.repr|) and optionally
a static part (of type |('c,int->int) R.repr|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)1$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.
